---
title: "Problem Set 3 - Statistics"
format: html
editor: visual
---

## Problem Set 3: Bats

Normal set up code:

```{r}
rm(list = ls())
library(tidyverse)
library(here)
library(ggfortify)
library(multcomp)
```

Read in and examine the data

```{r}
bats <- read.csv(here("Data", "bat_data.csv"))

glimpse(bats)
```

Looks like we have 9 variables and 9 variables. There are 5 character variables and 5 numeric/integer variables. Based on the questions we're asking, I will change 3 of the character variables into factor variables.

Make Species, age, sex into factor variables

```{r}
bats$Species <- as.factor(bats$Species)
levels(bats$Species)  #theres no typos in the data, we have the correct number of levels 

```

```{r}
bats$age <- as.factor(bats$age)
levels(bats$age)  #We're supposed to have two levels
```

```{r}
bats$sex <- as.factor(bats$sex)
levels(bats$sex)
```

Look at numeric data for outliers

```{r}
ggplot(bats, aes(x = mass_g, y = body_length_mm)) + 
  geom_point()

ggplot(bats, aes(x = tick_num, y = tail_length_mm)) + 
  geom_point()
```

Doesn't look like outliers will be an issue!

Start Analysis:

A. Does body mass differ between these five species of bats and, if so, how does body mass differ between the species?

For this question, I would run an ANOVA because we have a ratio response variable and a categorical predictor variable, and we're looking for a difference in means between groups.

**Build and Visualize the Model:**

```{r}
bats_modA <- lm(mass_g ~ Species, data = bats)

ggplot(bats, aes(x = Species, y = mass_g)) + 
  geom_boxplot() + 
  theme_bw()
```

Based on the boxplot, I would expect that the body mass for EPFU and LACI would be higher than the rest.

**Check Assumptions**

```{r}
autoplot(bats_modA, smooth.colour = NA)
```

Looks okay

**Make Summary and Anova Table**

```{r}
summary(bats_modA)

anova(bats_modA)
```

Based on the tables, there is a significant difference between the body mass for different species (F = 792.9, p = \<2.2e-16). Now test to see which are different.

**Tukey HSD Post-Hoc Test:**

```{r}
phcA <- glht(bats_modA, linfct = mcp(Species = "Tukey"))
summary(phcA)
```

```{r}
cld(phcA)
```

Looks like all of the species have significantly different mass from the other groups (p = \<1e-10).

**Make a more polished figure representing results:**

```{r}
ggplot(bats, aes(x = Species, y = mass_g, color = Species)) + 
  geom_point(size = 10, shape = "diamond", alpha = 0.5) + 
  coord_flip()+ 
  theme_bw()
```

B. Does body length differ between these five species of bats and, if so, how does body length differ between the species?

For this question, I would run an ANOVA because we have a ratio response variable and a categorical predictor variable, and we're looking for a difference in means between groups.

**Build and Visualize the model:**

```{r}
bats_modB <- lm(body_length_mm ~ Species, data = bats)

ggplot(bats, aes(x = Species, y = body_length_mm)) + 
  geom_boxplot() + 
  theme_bw()
```

Based on the boxplot, I would expect that all of the species would have significantly different body length from the other species.

**Check Assumptions**

```{r}
autoplot(bats_modB, smooth.colour = NA)
```

Looks good!

**Summary and ANOVA table**

```{r}
summary(bats_modB)

anova(bats_modB)
```

Based on the tables and p-values, we do have significant difference in body length between species (F = 70.498, p = \< 2.2e-16)

**Tukey HSD Post-Hoc Test**

```{r}
phcB <- glht(bats_modB, linfct = mcp(Species = "Tukey"))
summary(phcB)

cld(phcB)
```

The only groups that do not have statistically significantly difference in body length are MYLU and PESU (p = 0.26222), the rest are significantly different from each other.

**Make a more polished model representing the results:**

```{r}
ggplot(bats, aes(x = Species, y = body_length_mm, color = Species)) + 
  geom_point(size = 10, alpha = 0.5) + 
  coord_flip()+ 
  theme_bw()
```

C. Is the number of ticks found on the bats associated with their sex or age?

Because we have two categorical predictor variables and a discrete response variable, I would run a chi-squared test.

**Start by making a new table with the totals for each group**

```{r}
totals <- bats |>
  group_by(sex, age) |>
  summarise(total.number = sum(tick_num))

# Now make a bar graph becasue we have count data
ggplot(totals, aes(x = sex, y = total.number, fill = age))+
     geom_bar(stat = "identity", position = "dodge") + 
  theme_bw()
```

Looks like more ticks were found on males regardless of age based on the graph.

**Check assumptions**

We do not have a way to graphically check the assumptions for a chi-squared test. we just assume that the sampling was conducted randomly and that the observations are independent of each other.

**Make a matrix with the totals to run the chi-squared test**

```{r}
bats.mat <- xtabs(tick_num ~ sex + age, data = bats)
bats.mat
```

**Run the chi-squared test**

```{r}
chisq.test(bats.mat)
```

There is no statistical evidence that there is a difference in the number of ticks found on bats based on their age or sex (df = 1, p = 0.5615).

**Make a more formal plot**

```{r}
ggplot(totals, aes(x = sex, y = total.number, fill = age)) +
     geom_bar(stat = "identity", position = "dodge") + 
  theme_bw() + 
  xlab("Sex")+ 
  ylab("Number of Ticks") +
  scale_fill_manual(values = c(adult = "purple", subadult = "yellow"))
```

D. Disregarding species, is there a relationship in bats between tail length and body length?

We're testing for a relationship between two discrete variable, so I will run a linear regression.

**Make a scatter plot to visualize the data**

```{r}
ggplot(bats, aes(x = body_length_mm, y = tail_length_mm)) + 
  geom_point() + 
  theme_bw()
```

Looks like there is a positive slope, Intercept estimate: 14, slope estimate: 91/120 = 0.76


**Make the model**

```{r}
bats_modD <- lm(tail_length_mm ~ body_length_mm, data = bats)
```

**Check the assumptions**

```{r}
autoplot(bats_modD, smooth.colour = NA)
```

I don't see any issues with the assumptions

**Look at summary table**

```{r}
summary(bats_modD)
```

There is a statistically significant relationship between tail length and body length (F = 1037, df = 1 and 23, p = 2.2e-16).

**Make a more formal plot**

```{r}
ggplot(bats, aes(x = body_length_mm, y = tail_length_mm)) + 
  geom_point() + 
  geom_smooth(method = 'lm') +
  ylab("Tail Length (mm)") + 
  xlab("Body Length (mm)")
  theme_bw()
```
